---
title: "Create a sparse matrix from EHR data"
author: "Steve Simon"
date: "6/5/2018"
output: html_document
---

This program shows how to take information in the electronic health record and convert it to a sparse matrix format.

Please review open_oracle.Rmd to review how to access information in an Oracle database.

```{r timing-functions}
#
# This timing function is for my benefit. I want
# to informally check speeds and by writing the
# results to a text file, I can peek at the
# timing of early results while later results
# are still running. If you want to use this
# function, place log_time(first=TRUE) ins the
# first chunk to be timed and place log_time()
# in the remaining chunks.
#
program_name <- "extract_diabetes_patient_numbers"
timing_file <- paste0("../src/timing_", program_name, ".txt")
image_file <- paste0("../data/", program_name, ".RData")

log_time <- function(first=FALSE) {
  current_chunk <- opts_current$get("label")
  current_time <- as.character(Sys.time())
  cat(current_time, current_chunk, "began.\n", file=timing_file, append=!first)
}
```

```{r login-to-oracle}
log_time(first=TRUE)
library("dplyr")
library("magrittr")
library("Matrix")
library("ROracle")
cdm_config <- read.csv('../cdm_config_B2.csv', stringsAsFactors=FALSE)
c_connect <-
  dbConnect(Oracle(), cdm_config$account, cdm_config$password, cdm_config$access)
```

Run a simple test to see that the connection works. 

```{r simple-test}
log_time()
dbGetQuery(c_connect, "SELECT * FROM blueherondata.observation_fact WHERE rownum < 5")
dbGetQuery(c_connect, "SELECT * FROM blueherondata.concept_dimension WHERE rownum < 5")
dbGetQuery(c_connect, "SELECT * FROM blueheronmetadata.heron_terms WHERE rownum < 5")
```

Start small.

```{r medication-concepts}
log_time()
sql_medication_concepts <- "
SELECT DISTINCT concept_cd FROM blueherondata.concept_dimension
WHERE concept_path LIKE '\\i2b2\\Medications\\%'
ORDER BY concept_cd
"
db_medication_concepts <- dbGetQuery(c_connect, sql_medication_concepts)
sample_n(db_medication_concepts, size=10)
dim(db_medication_concepts)
```


```{r diabetes-patients, eval=FALSE}
log_time()
sql_patients_and_concepts <- "
SELECT patient_num concept_cd
FROM blueherondata.observation_fact
WHERE patient_num IN (
  SELECT patient_num, MIN(start_date) AS pdate
  FROM blueherondata.observation_fact
  WHERE concept_cd IN (
    SELECT DISTINCT c_basecode 
    FROM blueheronmetadata.heron_terms
    WHERE
      c_tooltip LIKE '%250 Diabetes mellitus%' OR
      c_tooltip LIKE '%E08-E13 Diabetes mellitus%'
     )
  GROUP BY patient_num
  ORDER BY patient_num
)
AND concept_cd IN (
  SELECT DISTINCT concept_cd FROM blueherondata.concept_dimension
  WHERE concept_path LIKE '\\i2b2\\Medications\\%'
  ORDER BY concept_cd
)"

sql_alternate_count <- "
SELECT patient_num concept_cd
FROM blueherondata.observation_fact
WHERE concept_cd IN (
  SELECT DISTINCT concept_cd FROM blueherondata.concept_dimension
  WHERE concept_path LIKE '\\i2b2\\Medications\\%'
)
AND patient_num IN (
  SELECT DISTINCT patient_num
  FROM blueherondata.observation_fact
  WHERE concept_cd IN (
    SELECT DISTINCT c_basecode 
    FROM blueheronmetadata.heron_terms
    WHERE
      c_tooltip LIKE '%250 Diabetes mellitus%' OR
      c_tooltip LIKE '%E08-E13 Diabetes mellitus%'
  )
)
"

diabetic_medications <- dbGetQuery(c_connect, sql_alternate_count)
dim(diabetic_medications)
save(diabetic_medications, file="diabetic_medications.RData")
```

Get counts.

```{r distinct-counts}
load(file="diabetic_medications.RData")
diabetic_medications %>%
  select(patient_num) %>%
  distinct %>%
  nrow -> n_patients 

diabetic_medications %>%
  select(concept_cd) %>%
  distinct %>%
  nrow -> n_concepts 

diabetic_medications %>%
  distinct %>%
  nrow -> n_patient_concepts

matrix_size <- as.numeric(n_patients)*as.numeric(n_concepts)
matrix_ratio <- 100*n_patient_concepts/matrix_size
matrix_ratio %<>% round(2)
matrix_ratio %<>% paste0("%")
```

Note that there are `r n_patients %>% format(big.mark=",")` distinct patients and `r n_concepts %>% format(big.mark=",")` distinct medications. If you arranged this in a traditional matrix with one row per patient and one indicator variable for every medication, you would have a matrix with `r matrix_size %>% format(big.mark=",")` entries. But there are only `r n_patient_concepts %>% format(big.mark=",")` unique entries in this database, meaning that only `r matrix_ratio` of the entries would be non-zero.

The sparse matrix format saves space and computation time by storing only the non-zero entries of the matrix. The details are a bit tricky. I tried to outline them in [a post on my blog](http://blog.pmean.com/sparse-matrices/).

The sparseMatrix function in the Matrix package will create a sparse matrix from a data frame with patient numbers and concept codes.

There are two storage options. The TsparseMatrix format stores the row location and the column location of the non-zero entries only. The CsparseMatrix format stores the row location and pointers. Normally the CsparseMatrix format uses less storage, but not always. Both formats are usually quite good for electronic health record data.

```{r create-sparse-matrix}
i <- factor(diabetic_medications$patient_num)
j <- factor(diabetic_medications$concept_cd)
sparse_c_format <- sparseMatrix(
  i=as.numeric(i),
  j=as.numeric(j),
  x=1,
  dimnames=list(levels(i), levels(j))
)
str(sparse_c_format)
size_c_format <- object.size(sparse_c_format)
efficiency_c_format <- 8*as.numeric(n_patients)*as.numeric(n_concepts)/size_c_format
sparse_t_format <- sparseMatrix(
  i=as.numeric(i),
  j=as.numeric(j),
  x=1,
  dimnames=list(levels(i), levels(j)),
  giveCsparse=FALSE
)
str(sparse_t_format)
size_t_format <- object.size(sparse_t_format)
efficiency_t_format <- 8*as.numeric(n_patients)*as.numeric(n_concepts)/size_t_format
sparse_binary <- sparseMatrix(
  i=as.numeric(i),
  j=as.numeric(j),
  dimnames=list(levels(i), levels(j)),
  giveCsparse=TRUE
)
str(sparse_binary)
size_binary_format <- object.size(sparse_binary)
efficiency_binary_format <- 8*as.numeric(n_patients)*as.numeric(n_concepts)/size_binary_format
```

The C format (which uses pointers) has a size of `r size_c_format %>% as.numeric %>% format(big.mark=",")` bytes compared to approximately `r (8*matrix_size) %>% format(big.mark=",")` bytes for a regular matrix. This is a `r round(efficiency_c_format)` fold improvement.

The T format (which uses row/column locations) has a size of `r size_t_format %>% as.numeric %>% format(big.mark=",")` bytes and represents a `r round(efficiency_t_format)` fold improvement.

The binary format (which takes advantage of the 0-1 coding) has a size of `r size_binary_format %>% as.numeric %>% format(big.mark=",")` bytes and represents a `r round(efficiency_binary_format)` fold improvement.

You can save a bit more space if you have a symmetric matrix, an upper or lower triangular matrix, a diagonal matrix, or a matrix with only zeros and ones. The zero/one format is called a binary or logical matrix.

Save everything for later use.

```{r save-everything}
log_time()
read.table(timing_file) 
save.image(image_file)
```