---
title: "Explore the EHR hierachy"
author: "Steve Simon"
date: "5/31/2018"
output: html_document
---

This program shows how to drill down through various levels of the hierarchy in the electronic health record.

Please review open_oracle.Rmd to review how to access information in an Oracle database.


```{r login-to-oracle}
library("ROracle")
cdm_config <- read.csv('../cdm_config.csv', stringsAsFactors=FALSE)
c_connect <-
  dbConnect(Oracle(), cdm_config$account, cdm_config$password, cdm_config$access)
```

Run a simple test to see that the connection works. 

```{r simple-test}
dbGetQuery(c_connect, "SELECT * FROM blueherondata.observation_fact WHERE rownum < 5")
dbGetQuery(c_connect, "SELECT * FROM blueherondata.concept_dimension WHERE rownum < 5")
dbGetQuery(c_connect, "SELECT * FROM blueheronmetadata.heron_terms WHERE rownum < 5")
```

The hierarchical structure of the electronic health record is represented by the concept path, a field in the concept_dimension table. There are similar representations of the hierachical structure in the heron_terms table of the blueheronmetadata database. These include the fields c_fullname, c_dimcode, and c_tooltip.

```{r count-paths}
sql_count_paths <- "
  SELECT COUNT(concept_path) 
    FROM blueherondata.concept_dimension"

sql_count_i2b2_paths <- "
  SELECT COUNT(concept_path) 
    FROM blueherondata.concept_dimension 
    WHERE concept_path LIKE '\\i2b2\\%'"

sql_count_medication_paths <- "
  SELECT COUNT(concept_path) 
    FROM blueherondata.concept_dimension 
    WHERE concept_path LIKE '\\i2b2\\Medications\\%'"

dbGetQuery(c_connect, sql_count_paths)
dbGetQuery(c_connect, sql_count_i2b2_paths)
dbGetQuery(c_connect, sql_count_medication_paths)
```

The information in the blueheronmetadata database is fairly similar.

```{r count-metadata}
sql_count_metadata_paths <- "
  SELECT COUNT(c_fullname) 
    FROM blueheronmetadata.heron_terms"

sql_count_metadata_i2b2_paths <- "
  SELECT COUNT(c_fullname) 
    FROM blueheronmetadata.heron_terms
    WHERE c_fullname LIKE '\\i2b2\\%'"

sql_count_metadata_medication_paths <- "
  SELECT COUNT(c_fullname) 
    FROM blueheronmetadata.heron_terms
    WHERE c_fullname LIKE '\\i2b2\\Medications\\%'"

dbGetQuery(c_connect, sql_count_metadata_paths)
dbGetQuery(c_connect, sql_count_metadata_i2b2_paths)
dbGetQuery(c_connect, sql_count_metadata_medication_paths)
```

Note that c_fullname and c_dimcode are almost identical, at least in this particular database.

```{r identical}
sql_count_discrepancies <- "
  SELECT COUNT(c_fullname) 
    FROM blueheronmetadata.heron_terms
    WHERE c_fullname <> c_dimcode"

sql_count_i2b2_discrepancies <- "
  SELECT COUNT(c_fullname) 
    FROM blueheronmetadata.heron_terms
    WHERE
      c_fullname <> c_dimcode AND
      c_fullname LIKE '\\i2b2\\%'"

dbGetQuery(c_connect, sql_count_discrepancies)
dbGetQuery(c_connect, sql_count_i2b2_discrepancies)
```

Let's explore the hierarchy for medications, as that is a reasonably manageable size.

```{r medications}
sql_extract_medication_paths <- "
  SELECT concept_path, concept_cd, name_char 
    FROM blueherondata.concept_dimension
    WHERE concept_path LIKE '\\i2b2\\Medications\\%'"

sql_extract_metadata_medication_paths <- "
  SELECT c_fullname, c_tooltip, c_basecode, c_name 
    FROM blueheronmetadata.heron_terms
    WHERE c_fullname LIKE '\\i2b2\\Medications\\%'"

medication_paths <- dbGetQuery(c_connect, sql_extract_medication_paths)
medication_metadata <- dbGetQuery(c_connect, sql_extract_metadata_medication_paths)

names(medication_paths) <- tolower(names(medication_paths))
names(medication_metadata) <- tolower(names(medication_metadata))
```

There's a very close correspondance between concept_path and c_fullname.

```{r concept}
setdiff(medication_paths$concept_path, medication_metadata$c_fullname)
setdiff(medication_metadata$c_fullname, medication_paths$concept_path)
```

and for concept_cd and c_basecode

```{r cd}
setdiff(medication_paths$concept_cd, medication_metadata$c_basecode)
setdiff(medication_metadata$c_basecode, medication_paths$concept_cd)
```

but name_char and c_name don't seem to line up as nicely.

```{r name}
library(dplyr)
library(magrittr)
length(setdiff(medication_paths$name_char, medication_metadata$c_name))
length(setdiff(medication_metadata$c_name, medication_paths$name_char))

medication_paths %>%
  inner_join(medication_metadata, by=c("concept_path"="c_fullname")) %>%
  select(name_char, c_name) %>%
  filter(name_char != c_name) %>%
  sample_n(size=100) %>%
  mutate(x=paste(name_char, c_name, sep="||")) %>%
  select(x)

medication_metadata$c_name <- sub(" \\[.*\\]$", "", medication_metadata$c_name)

length(setdiff(medication_paths$name_char, medication_metadata$c_name))
length(setdiff(medication_metadata$c_name, medication_paths$name_char))
```