---
title: "Extract ids of patients with diabetes"
author: "Steve Simon"
date: "5/31/2018"
output: html_document
---

This program shows how to extract the patient numbers for diabetes patients subject to a variety of restrictions.

Load the relevant libraries.

```{r libraries}
library(dplyr)
library(knitr)
library(magrittr)
library("ROracle")
```

Please review open_oracle.Rmd to review how to access information in an Oracle database.

```{r timing-functions}
program_name <- "extract_diabetes_patient_numbers"
timing_file <- paste0("../src/", program_name, "_timing.txt")
image_file <- paste0("../data/", program_name, ".RData")

log_time <- function(first=FALSE) {
  current_chunk <- opts_current$get("label")
  current_time <- as.character(Sys.time())
  cat(current_time, current_chunk, "began.\n", file=timing_file, append=!first)
}
```


```{r login-to-oracle}
log_time(first=TRUE)
cdm_config <- read.csv('../cdm_config.csv', stringsAsFactors=FALSE)
c_connect <-
  dbConnect(Oracle(), cdm_config$account, cdm_config$password, cdm_config$access)
```


The SQL code to count and extract patient numbers can get quite long. It helps to break the code into distinct pieces.

The following SQL snippet counts the distinct number of patients. You can combine with additional SQL code to place restrictions on the types of patients.

```{r count}
log_time()
sql_count_patients <- "
SELECT COUNT(DISTINCT patient_num)
FROM blueherondata.observation_fact
"

dbGetQuery(c_connect, sql_count_patients)
```

To make the SQL code look nicer, here is a function that indents a piece of SQL code nested between two other pieces.

```{r nest}
nest <- function(sql1, sql2, sql3="") {
  paste0(
    sql1,
    gsub("\n", "\n  ", sql2),
    sql3)
}
```

This SQL snippet adds a restriction to the ICD9 and ICD10 codes associated with diabetes.

```{r diabetes-1}
log_time()

sql_select_diabetes_codes <- "
SELECT DISTINCT concept_cd 
FROM blueherondata.concept_dimension
WHERE
  concept_cd LIKE 'ICD9:250%' OR
  concept_cd LIKE 'ICD10:E08%' OR
  concept_cd LIKE 'ICD10:E09%' OR
  concept_cd LIKE 'ICD10:E10%' OR
  concept_cd LIKE 'ICD10:E11%' OR
  concept_cd LIKE 'ICD10:E12%' OR
  concept_cd LIKE 'ICD10:E13%'
"

dbGetQuery(c_connect, sql_select_diabetes_codes)

sql_restrict_to_diabetes <- nest("WHERE concept_cd IN (", sql_select_diabetes_codes, ")")
cat(sql_restrict_to_diabetes)
```

Combine this with the previous string.

```{r diabetes-2}
log_time()
sql_count_diabetics <- nest(sql_count_patients, sql_restrict_to_diabetes)
cat(sql_count_diabetics)
dbGetQuery(c_connect, sql_count_diabetics)
```

This snippet adds a date restriction.

```{r dates-1}
log_time()
sql_restrict_dates <- "
AND TO_CHAR(start_date, 'yyyy-mm-dd') > '2010-01-01'
"
```

Here's what the combination of all three strings looks like.

```{r dates-2}
log_time()
sql_count_date_restricted <- 
  nest(sql_count_diabetics, sql_restrict_dates)
cat(sql_count_date_restricted)
dbGetQuery(c_connect, sql_count_date_restricted)
```

This snippet restricts to patients to females only. This represents a self-join.

```{r females-1}
log_time()
sql_restrict_to_females <- "
AND patient_num IN (
  SELECT DISTINCT patient_num
  FROM blueherondata.observation_fact
  WHERE
    concept_cd = 'DEM|SEX:f'
)"
sql_count_female_diabetics <- nest(sql_count_date_restricted, sql_restrict_to_females)
cat(sql_count_female_diabetics)
```

Unfortunately, it is rather slow, so I do not recommend that you use this approach.

```{r females-2, eval=FALSE}
log_time()
dbGetQuery(c_connect, sql_combine)
```

Extracting all the patient numbers requires a different start to the SQL query.

```{r extract-1}
log_time()
sql_extract_patients <- "
SELECT patient_num, start_date AS pdate
  FROM blueherondata.observation_fact
"
```

This is what the combined query looks like.

```{r extract-2}
log_time()
sql_extract_diabetics <- nest(sql_extract_patients, sql_restrict_to_diabetes)
cat(sql_extract_diabetics)

patient_list <- dbGetQuery(c_connect, sql_extract_diabetics)
names(patient_list) <- tolower(names(patient_list))
patient_list %<>% arrange(patient_num)
head(patient_list)
dim(patient_list)
```

Notice that patients appear more often than once in the query. To get a distinct set of patients, you need to group by patient number and then choose only one record. The simplest choice is the record with the smallest start-date.

you can do this within R, but be careful, because some of the base R functions that you'd be tempted to use, such as unique or duplicated can be very inefficient with large data frames. The distinct or summarize functions in dplyr are a better choice.

```{r distinct-1}
log_time()
library(dplyr)
library(magrittr)
patient_list                   %>% 
  group_by(patient_num)        %>%
  summarize(pdate=min(pdate))  %>%
  ungroup                      -> unique_patients_1
head(unique_patients_1)
dim(unique_patients_1)
```

This method also works.

```{r distinct-2}
log_time()
library(dplyr)
library(magrittr)
patient_list                             %>% 
  arrange(patient_num, pdate)            %>%
  distinct(patient_num, .keep_all=TRUE)  %>%
  ungroup                                -> unique_patients_2
head(unique_patients_2)
dim(unique_patients_2)
```

You can also get the same effect using just SQL code.

```{r distinct-3}
log_time()
sql_extract_unique_patients <- "
SELECT patient_num, MIN(start_date) AS pdate
FROM blueherondata.observation_fact
"

sql_group <- "
GROUP BY patient_num
ORDER BY patient_num"
```

```{r distinct-4}
log_time()
sql_extract_unique_diabetics <- nest(sql_extract_unique_patients, sql_restrict_to_diabetes, sql_group)
cat(sql_extract_unique_diabetics)
unique_patients_3 <- dbGetQuery(c_connect, sql_extract_unique_diabetics)
head(unique_patients_3)
dim(unique_patients_3)
```

Now, let's take things one step further. For these unique patients, let's pull any record associated with medications.

```{r medication}
sql_extract_medications_prefix <- "
SELECT patient_num, concept_cd, start_date AS edate
FROM blueherondata.observation_fact

WHERE patient_num IN ("
  
sql_extract_medications_postfix <- "
) AND 

concept_cd IN (
SELECT DISTINCT concept_cd FROM blueherondata.concept_dimension
  WHERE concept_path LIKE '\\i2b2\\Medications\\%'
  ORDER BY concept_cd)"

sql_extract_medications <- paste(
  sql_extract_medications_prefix,
  sql_extract_unique_patients,
  sql_restrict_to_diabetes,
  sql_group,
  sql_extract_medications_postfix)
cat(sql_extract_medications)
```

Save everything for possible re-use

```{r save-everything}
log_time()
read.table(timing_file) 
save.image(image_file)
```
