---
title: "Extract ids of patients with diabetes"
author: "Steve Simon"
date: "5/31/2018"
output: html_document
---

This program shows how to extract the patient numbers for diabetes patients subject to a variety of restrictions.

Please review open_oracle.Rmd to review how to access information in an Oracle database.

```{r timing-functions}
program_name <- "extract_diabetes_patient_numbers"
timing_file <- paste0("../src/", program_name, "_timing.txt")
image_file <- paste0("../data/", program_name, ".RData")

library(knitr)
log_time <- function(first=FALSE) {
  current_chunk <- opts_current$get("label")
  current_time <- as.character(Sys.time())
  cat(current_time, current_chunk, "began.\n", file=timing_file, append=!first)
}
```


```{r login-to-oracle}
log_time(first=TRUE)
library("ROracle")
cdm_config <- read.csv('../cdm_config.csv', stringsAsFactors=FALSE)
c_connect <-
  dbConnect(Oracle(), cdm_config$account, cdm_config$password, cdm_config$access)
```

Run a simple test to see that the connection works. 

```{r simple-test}
log_time()
dbGetQuery(c_connect, "select count(patient_num) from blueherondata.observation_fact")
```
```

The SQL code to count and extract patient numbers can get quite long. It helps to break the code into distinct pieces.

The following SQL snippet counts the distinct number of patients. You can combine with additional SQL code to place restrictions on the types of patients.

```{r count}
log_time()
sql_count_patients <- "
  SELECT COUNT(DISTINCT patient_num)
    FROM blueherondata.observation_fact
"

dbGetQuery(c_connect, sql_count_patients)
```

This SQL snippet adds a restriction to the ICD9 and ICD10 codes associated with diabetes.

```{r diabetes-1}
log_time()
sql_restrict_to_diabetes <- "
  WHERE concept_cd IN (
    SELECT DISTINCT concept_cd 
    FROM blueherondata.concept_dimension
    WHERE
      concept_cd LIKE 'ICD9:250%' OR
      concept_cd LIKE 'ICD10:E08%' OR
      concept_cd LIKE 'ICD10:E09%' OR
      concept_cd LIKE 'ICD10:E10%' OR
      concept_cd LIKE 'ICD10:E11%' OR
      concept_cd LIKE 'ICD10:E12%' OR
      concept_cd LIKE 'ICD10:E13%'
  )
"
```

Combine this with the previous string.

```{r diabetes-2}
log_time()
sql_combine <- paste(
  sql_count_patients, 
  sql_restrict_to_diabetes)
cat(sql_combine)
dbGetQuery(c_connect, sql_combine)
```

This snippet adds a date restriction.

```{r dates-1}
log_time()
sql_restrict_dates <- "
  AND TO_CHAR(start_date, 'yyyy-mm-dd') > '2010-01-01'
"
```

Here's what the combination of all three strings looks like.

```{r dates-2}
log_time()
sql_combine <- paste(
  sql_count_patients, 
  sql_restrict_to_diabetes,
  sql_restrict_dates)
cat(sql_combine)
dbGetQuery(c_connect, sql_combine)
```

This snippet restricts to patients to females only. This represents a self-join.

```{r females-1}
log_time()
sql_restrict_to_females <- "
  AND patient_num IN (
    SELECT DISTINCT patient_num
    FROM blueherondata.observation_fact
    WHERE
      concept_cd = 'DEM|SEX:f'
  )
"
```


```{r females-2}
log_time()
sql_combine <- paste(
  sql_count_patients, 
  sql_restrict_to_diabetes,
  sql_restrict_dates,
  sql_restrict_to_females)
cat(sql_combine)
dbGetQuery(c_connect, sql_combine)
```

Extracting all the patient numbers requires a different start to the SQL query.

```{r extract-1}
log_time()
sql_extract_patients <- "
  SELECT patient_num, start_date AS pdate
    FROM blueherondata.observation_fact
"
```

This is what the combined query looks like.

```{r extract-2}
log_time()
sql_combine <- paste(
  sql_extract_patients,
  sql_restrict_to_diabetes)
cat(sql_combine)

patient_list <- dbGetQuery(c_connect, sql_combine)
names(patient_list) <- tolower(names(patient_list))
head(patient_list)
dim(patient_list)
```

Notice that patients appear more often than once in the query. To get a distinct set of patients, you need to group by patient number and then choose only one record. The simplest choice is the record with the smallest start-date.

you can do this within R, but be careful, because some of the base R functions that you'd be tempted to use, such as unique or duplicated can be very inefficient with large data frames. The distinct or summarize functions in dplyr are a better choice.

```{r distinct-1}
log_time()
library(dplyr)
library(magrittr)
patient_list                   %>% 
  group_by(patient_num)        %>%
  summarize(pdate=min(pdate))  %>%
  ungroup                      -> unique_patients_1
head(unique_patients_1)
dim(unique_patients_1)
```

This method also works.

```{r distinct-2}
log_time()
library(dplyr)
library(magrittr)
patient_list                             %>% 
  arrange(patient_num, pdate)            %>%
  distinct(patient_num, .keep_all=TRUE)  %>%
  ungroup                                -> unique_patients_2
head(unique_patients_2)
dim(unique_patients_2)
```

You can also get the same effect using just SQL code.

```{r distinct-3}
log_time()
sql_extract_unique_patients <- "
  SELECT patient_num, MIN(start_date) AS pdate
    FROM blueherondata.observation_fact
"

sql_group <- "
  GROUP BY patient_num
  ORDER BY patient_num"
```

```{r distinct-4}
log_time()
sql_combine <- paste(
  sql_extract_unique_patients,
  sql_restrict_to_diabetes,
  sql_group)
cat(sql_combine)
unique_patients_3 <- dbGetQuery(c_connect, sql_combine)
head(unique_patients_3)
dim(unique_patients_3)
```

Save everything for possible re-sue

```{r save-everything}
log_time()
read.table(timing_file) 
save.image(image_file)
```
