---
title: "Apply exclusions to a patient list"
author: "Steve Simon"
date: "5/31/2018"
output: html_document
---

This program shows how to apply exclusions to a patient list, using diabetes as an example.

Load the relevant libraries. 

```{r libraries}
library(dplyr)
library(knitr)
library(magrittr)
library(ROracle)
```

Please review open_oracle.Rmd to review how to access information in an Oracle database.

```{r timing-functions}
#
# This timing function is for my benefit. I want
# to informally check speeds and by writing the
# results to a text file, I can peek at the
# timing of early results while later results
# are still running. If you want to use this
# function, place log_time(first=TRUE) ins the
# first chunk to be timed and place log_time()
# in the remaining chunks.
#
program_name <- "extract_diabetes_patient_numbers"
timing_file <- paste0("../src/timing_", program_name, ".txt")
image_file <- paste0("../data/", program_name, ".RData")

log_time <- function(first=FALSE) {
  current_chunk <- opts_current$get("label")
  current_time <- as.character(Sys.time())
  cat(current_time, current_chunk, "began.\n", file=timing_file, append=!first)
}
```


```{r login-to-oracle}
log_time(first=TRUE)
cdm_config <- read.csv('../cdm_config_B2.csv', stringsAsFactors=FALSE)
c_connect <-
  dbConnect(Oracle(), cdm_config$account, cdm_config$password, cdm_config$access)
```

Quick test

```{r quick-test}
dbGetQuery(c_connect, "SELECT * FROM blueherondata.observation_fact WHERE ROWNUM < 5")
dbGetQuery(c_connect, "SELECT * FROM blueherondata.concept_dimension WHERE ROWNUM < 5")
dbGetQuery(c_connect, "SELECT * FROM blueheronmetadata.heron_terms WHERE ROWNUM < 5")
```


The SQL code to count and extract patient numbers can get quite long. It helps to build the code in stages.

The following SQL snippet counts the distinct number of patients. You can combine with additional SQL code to place restrictions on the types of patients.

```{r count}
log_time()
sql_count_patients <- "
SELECT COUNT(DISTINCT patient_num)
FROM blueherondata.observation_fact
"

dbGetQuery(c_connect, sql_count_patients) %>%
  format(big.mark=",")
```

To make the SQL code look nicer, here is a function that indents a piece of SQL code nested between two other pieces. I need to make the transition to a more formal package like glue, which includes a glue_sql function. This function makes it easy to use nested queries and protects against SQL injection attacks by using a parameterized SQL query. The glue package is part of the Tidyverse and has a [nice overview page](https://glue.tidyverse.org/).

```{r nest}
nest <- function(sql1, sql2, sql3="") {
  paste0(
    sql1,
    gsub("\n", "\n  ", sql2),
    sql3)
}
nestx <- function(sql1, sql2, sql3="", sql4="") {
  paste0(
    sql1,
    gsub("\n", "\n  ", sql2),
    gsub("\n", "\n    ", sql3),
    sql4)
}
```

This SQL snippet lists all the ICD9 and ICD10 codes associated with diabetes. A sample of the codes is shown.

First the ICD9 codes.

```{r icd9-diabetes-codes}
log_time()

sql_list_diabetes_icd9 <- "
SELECT DISTINCT concept_cd 
FROM blueherondata.concept_dimension
WHERE concept_cd LIKE '%ICD9:250%'
"

diabetes_codes <- dbGetQuery(c_connect, sql_list_diabetes_icd9)
diabetes_codes %>%
  dim
diabetes_codes %>%
  sample_n(5) %>%
  unlist %>%
  sort %>%
  cat(sep="\n")



```

Here are the ICD10 codes.

```{r icd10-diabetes-codes}
sql_list_diabetes_icd10 <- "
SELECT DISTINCT concept_cd 
FROM blueherondata.concept_dimension
WHERE
  concept_cd LIKE '%ICD10:E08%' OR
  concept_cd LIKE '%ICD10:E09%' OR
  concept_cd LIKE '%ICD10:E10%' OR
  concept_cd LIKE '%ICD10:E11%' OR
  concept_cd LIKE '%ICD10:E12%' OR
  concept_cd LIKE '%ICD10:E13%'
"

diabetes_codes <- dbGetQuery(c_connect, sql_list_diabetes_icd10)
diabetes_codes %>%
  dim
diabetes_codes %>%
  sample_n(5) %>%
  unlist %>%
  sort %>%
  cat(sep="\n")
```

Combine these two queries together and nest them within a counting query.

```{r diabetes-count}
sql_count_diabetics <- "
SELECT COUNT(DISTINCT patient_num)
FROM blueherondata.observation_fact
WHERE concept_cd IN (
  SELECT DISTINCT concept_cd 
  FROM blueherondata.concept_dimension
  WHERE
    concept_cd LIKE '%ICD9:250%' OR
    concept_cd LIKE '%ICD10:E08%' OR
    concept_cd LIKE '%ICD10:E09%' OR
    concept_cd LIKE '%ICD10:E10%' OR
    concept_cd LIKE '%ICD10:E11%' OR
    concept_cd LIKE '%ICD10:E12%' OR
    concept_cd LIKE '%ICD10:E13%'
)
"

dbGetQuery(c_connect, sql_count_diabetics) %>%
  format(big.mark=",")
```

Information about diabetes codes is also available through the concept_path variable found in the concept_dimension table.

Here is a random sample of paths associated with ICD9 diabetes codes...

```{r explore-path-icd9}
log_time()

sql_explore_icd9_path <- "
SELECT DISTINCT concept_path
FROM blueherondata.concept_dimension
WHERE concept_cd LIKE '%ICD9:250%'
ORDER BY concept_path
"

paths <- dbGetQuery(c_connect, sql_explore_icd9_path)
paths %>%
  dim
paths %>%
  sample_n(10) %>%
  unlist %>%
  sort %>%
  cat(sep="\n")
```

...and ICD10 diabetes codes.

```{r explore-icd10-path}
sql_explore_icd10_path <- "
SELECT DISTINCT concept_path
FROM blueherondata.concept_dimension
WHERE
  concept_cd LIKE '%ICD10:E08%' OR
  concept_cd LIKE '%ICD10:E09%' OR
  concept_cd LIKE '%ICD10:E10%' OR
  concept_cd LIKE '%ICD10:E11%' OR
  concept_cd LIKE '%ICD10:E12%' OR
  concept_cd LIKE '%ICD10:E13%'
ORDER BY concept_path
"

paths <- dbGetQuery(c_connect, sql_explore_icd10_path)
paths %>%
  dim
paths %>%
  sample_n(5) %>%
  unlist %>%
  sort %>%
  cat(sep="\n")
```

Unfortunately, the concept_path variable is mostly coded using cryptic values. If you stare at this list long enough, you'll figure out that the codes "A8352361" or "(250) Diabetes mellitus" will give you the proper ICD9 codes and that "A18916268" will give you the proper ICD10 codes 

You can find cleaner descriptions using c_tooltip in blueheronmetadata.heron_terms. The halibut system uses c_tooltip for tool tip displays, so this value is almost always easier to read.

Here is a random sample of tooltips associated with ICD9 diabetes codes...

```{r explore-icd9-tooltip}
sql_explore_icd9_tooltip <- "
SELECT DISTINCT c_tooltip 
FROM blueheronmetadata.heron_terms
WHERE
  c_basecode LIKE '%ICD9:250%'
ORDER BY c_tooltip
"

tooltips <- dbGetQuery(c_connect, sql_explore_icd9_tooltip)
tooltips %>%
  dim
tooltips %>%
  sample_n(5) %>%
  unlist %>%
  sort %>%
  cat(sep="\n")
```

...and ICD10 diabetes codes.

```{r explore-icd10-tootip}
sql_explore_icd10_tooltip <- "
SELECT DISTINCT c_tooltip 
FROM blueheronmetadata.heron_terms
WHERE
  c_basecode LIKE '%ICD10:E08%' OR
  c_basecode LIKE '%ICD10:E09%' OR
  c_basecode LIKE '%ICD10:E10%' OR
  c_basecode LIKE '%ICD10:E11%' OR
  c_basecode LIKE '%ICD10:E12%' OR
  c_basecode LIKE '%ICD10:E13%'
ORDER BY c_tooltip
"

tooltips <- dbGetQuery(c_connect, sql_explore_icd10_tooltip)
tooltips %>%
  dim
tooltips %>%
  sample_n(5) %>%
  unlist %>%
  sort %>%
  cat(sep="\n")
```

Very nice! For ICD9 codes associated with diabetes, use "250 Diabetes mellitus"" and for ICD10 codes associated with diabetes, use "E08-E13 Diabetes mellitus". You can find key phrases in the tool tip using Heron as well.

Let's flip this around and find if there are any other concept_cd/c_basecode values associated with c_tooltip values containing diabetes codes.

First, let's look at the ICD9 side of things...

```{r alternate-codes-icd9}
log_time()
sql_alternate_icd9_codes <- "
SELECT DISTINCT c_basecode 
FROM blueheronmetadata.heron_terms
WHERE
  c_tooltip LIKE '%250 Diabetes mellitus%'
"

alternate_codes <- dbGetQuery(c_connect, sql_alternate_icd9_codes)
alternate_codes %>% 
  dim
alternate_codes %>%
  sample_n(5) %>%
  arrange(C_BASECODE)
```


.. and the ICD10 side.

```{r alternate-codes-icd10}
sql_alternate_icd10_codes <- "
SELECT DISTINCT c_basecode 
FROM blueheronmetadata.heron_terms
WHERE
  c_tooltip LIKE '%E08-E13 Diabetes mellitus%'
"

alternate_codes <- dbGetQuery(c_connect, sql_alternate_icd10_codes)
alternate_codes %>% 
  dim
alternate_codes %>%
  sample_n(5) %>%
  arrange(C_BASECODE)
```

What are the labels associated with all these codes?

```{r labels}
sql_labels <- "
SELECT DISTINCT c_name
FROM blueheronmetadata.heron_terms
WHERE
  c_tooltip LIKE '%250 Diabetes mellitus%'
  ORDER BY c_name
"
dbGetQuery(c_connect, sql_labels)

sql_labels <- "
SELECT DISTINCT c_name
FROM blueheronmetadata.heron_terms
WHERE
  c_tooltip LIKE '%E08-E13 Diabetes mellitus%'
  ORDER BY c_name
"
dbGetQuery(c_connect, sql_labels)
```


The c_basecode field in blueheronmetadata (which corresponds to the concept_cd field in blueherondata) has an interesting prefix, KUH|DX_ID which represents an internal diagnosis code.

If you want to get a larger set of diabetes patients, you need to take these internal diagnosis codes into account.

```{r alternate-count}
log_time()

sql_alternate_count <- "
SELECT COUNT(DISTINCT patient_num)
FROM blueherondata.observation_fact
WHERE concept_cd IN (
  SELECT DISTINCT c_basecode 
  FROM blueheronmetadata.heron_terms
  WHERE
    c_tooltip LIKE '%250 Diabetes mellitus%' OR
    c_tooltip LIKE '%E08-E13 Diabetes mellitus%'
)
"

dbGetQuery(c_connect, sql_alternate_count) %>%
  format(big.mark=",")
```

Extracting all the patient numbers requires a different start to the SQL query.

```{r extract-patients}
log_time()

sql_extract_diabetics <- "
SELECT patient_num, start_date AS pdate
FROM blueherondata.observation_fact
WHERE concept_cd IN (
  SELECT DISTINCT c_basecode 
  FROM blueheronmetadata.heron_terms
  WHERE
    c_tooltip LIKE '%250 Diabetes mellitus%' OR
    c_tooltip LIKE '%E08-E13 Diabetes mellitus%'
)
"

patient_list <- dbGetQuery(c_connect, sql_extract_diabetics)
names(patient_list) %<>% tolower
patient_list %>%
  dim
patient_list %>%
  sample_n(5) %>%
  arrange(patient_num)
```

Notice that patients appear more often than once in the query. To get a distinct set of patients, you need to group by patient number and then choose only one record. The simplest choice is the record with the earliest start-date.

you can do this within R, but be careful, because some of the base R functions that you'd be tempted to use, such as unique or duplicated can be very inefficient with large data frames. The distinct or summarize functions in dplyr are a better choice.

```{r extract-distinct}
log_time()
library(dplyr)
library(magrittr)
patient_list                   %>% 
  group_by(patient_num)        %>%
  summarize(pdate=min(pdate))  %>%
  ungroup                      -> unique_patients_1
unique_patients_1 %>%
  dim
unique_patients_1 %>%
  sample_n(5) %>%
  arrange(patient_num)
```

This method also works.

```{r alternate-distinct}
log_time()
patient_list                             %>% 
  arrange(patient_num, pdate)            %>%
  distinct(patient_num, .keep_all=TRUE)  %>%
  ungroup                                -> unique_patients_2
unique_patients_2 %>%
  dim
unique_patients_2 %>%
  sample_n(5) %>%
  arrange(patient_num)
```

You can also get the same effect using just SQL code.

```{r distinct-using-sql}
log_time()

sql_extract_unique_diabetics <- "
SELECT patient_num, MIN(start_date) AS pdate
FROM blueherondata.observation_fact
WHERE concept_cd IN (
  SELECT DISTINCT c_basecode 
  FROM blueheronmetadata.heron_terms
  WHERE
    c_tooltip LIKE '%250 Diabetes mellitus%' OR
    c_tooltip LIKE '%E08-E13 Diabetes mellitus%'
)
GROUP BY patient_num
ORDER BY patient_num
"

unique_diabetics <- dbGetQuery(c_connect, sql_extract_unique_diabetics)
names(unique_diabetics) %<>% tolower
unique_diabetics %>%
  dim
unique_diabetics %>%
  sample_n(5) %>%
  arrange(patient_num)
```

There is a fourth choice here. You can use a new package in the tidyverse called dbplyr. It takes many of the functions in the dplyr package and translates their data manipulations into SQL code. The dbplyr package is also part of the Tidyverse and also has a [nice overview page](https://dbplyr.tidyverse.org/). The current version of the dbplyr package does not yet support Oracle, but this support will be coming soon.

Save this SQL code and this list. You'll need it later.

```{r save_distinct}
save(list=c("sql_extract_unique_diabetics", "unique_diabetics"), file="../data/diabetes_patients.RData")
```

Save everything for possible re-use

```{r save-everything}
log_time()
read.table(timing_file) 
save.image(image_file)
```
