---
title: "Extract ids of patients with diabetes"
author: "Steve Simon"
date: "5/31/2018"
output: html_document
---

This program shows how to extract the patient numbers for diabetes patients subject to a variety of restrictions.

Load the relevant libraries.

```{r libraries}
library(dplyr)
library(knitr)
library(magrittr)
library("ROracle")
```

Please review open_oracle.Rmd to review how to access information in an Oracle database.

```{r timing-functions}
program_name <- "extract_diabetes_patient_numbers"
timing_file <- paste0("../src/", program_name, "_timing.txt")
image_file <- paste0("../data/", program_name, ".RData")

log_time <- function(first=FALSE) {
  current_chunk <- opts_current$get("label")
  current_time <- as.character(Sys.time())
  cat(current_time, current_chunk, "began.\n", file=timing_file, append=!first)
}
```


```{r login-to-oracle}
log_time(first=TRUE)
cdm_config <- read.csv('../cdm_config_A1.csv', stringsAsFactors=FALSE)
c_connect <-
  dbConnect(Oracle(), cdm_config$account, cdm_config$password, cdm_config$access)
```

Quick test

```{r quick-test}
dbGetQuery(c_connect, "SELECT * FROM blueherondata.observation_fact WHERE ROWNUM < 5")
dbGetQuery(c_connect, "SELECT * FROM blueherondata.concept_dimension WHERE ROWNUM < 5")
dbGetQuery(c_connect, "SELECT * FROM blueheronmetadata.heron_terms WHERE ROWNUM < 5")
```


The SQL code to count and extract patient numbers can get quite long. It helps to break the code into distinct pieces.

The following SQL snippet counts the distinct number of patients. You can combine with additional SQL code to place restrictions on the types of patients.

```{r count}
log_time()
sql_count_patients <- "
SELECT COUNT(DISTINCT patient_num)
FROM blueherondata.observation_fact
"

dbGetQuery(c_connect, sql_count_patients) %>%
  format(big.mark=",")
```

To make the SQL code look nicer, here is a function that indents a piece of SQL code nested between two other pieces.

```{r nest}
nest <- function(sql1, sql2, sql3="") {
  paste0(
    sql1,
    gsub("\n", "\n  ", sql2),
    sql3)
}
```

This SQL snippet lists all the ICD9 and ICD10 codes associated with diabetes.

```{r diabetes-codes}
log_time()

sql_list_diabetes_codes <- "
SELECT DISTINCT concept_cd 
FROM blueherondata.concept_dimension
WHERE
  concept_cd LIKE '%ICD9:250%' OR
  concept_cd LIKE '%ICD10:E08%' OR
  concept_cd LIKE '%ICD10:E09%' OR
  concept_cd LIKE '%ICD10:E10%' OR
  concept_cd LIKE '%ICD10:E11%' OR
  concept_cd LIKE '%ICD10:E12%' OR
  concept_cd LIKE '%ICD10:E13%'
"

diabetes_codes <- dbGetQuery(c_connect, sql_list_diabetes_codes)
diabetes_codes %>%
  dim
diabetes_codes %>%
  sample_n(20) %>%
  arrange(CONCEPT_CD)
diabetes_codes %>%
  unlist %>%
  sub(":.*$", "", .) %>%
  table

sql_select_diabetes <- nest("WHERE concept_cd IN (", sql_list_diabetes_codes, ")")
sql_count_diabetics <- nest(sql_count_patients, sql_select_diabetes)
cat(sql_count_diabetics)
dbGetQuery(c_connect, sql_count_diabetics) %>%
  format(big.mark=",")
```

Information about diabetes codes is also available through the concept_path variable found in the concept_dimension table.

```{r explore-path}
log_time()

sql_explore_concept_path <- "
SELECT DISTINCT concept_path
FROM blueherondata.concept_dimension
WHERE
  concept_cd LIKE '%ICD9:250%' OR
  concept_cd LIKE '%ICD10:E08%' OR
  concept_cd LIKE '%ICD10:E09%' OR
  concept_cd LIKE '%ICD10:E10%' OR
  concept_cd LIKE '%ICD10:E11%' OR
  concept_cd LIKE '%ICD10:E12%' OR
  concept_cd LIKE '%ICD10:E13%'
ORDER BY concept_path
"

paths <- dbGetQuery(c_connect, sql_explore_concept_path)
paths %>%
  unlist %>%
  cat(sep="\n")
```

Unfortunately, the concept_path variable is mostly coded using cryptic values. You can find cleaner descriptions using c_tooltip in blueheronmetadata.heron_terms. The halibut system uses c_tooltip for tool tip displays, so this value is almost always easier to read.

```{r explore-tooltip}
sql_explore_tooltip <- "
SELECT DISTINCT c_tooltip 
FROM blueheronmetadata.heron_terms
WHERE
  c_basecode LIKE '%ICD9:250%' OR
  c_basecode LIKE '%ICD10:E08%' OR
  c_basecode LIKE '%ICD10:E09%' OR
  c_basecode LIKE '%ICD10:E10%' OR
  c_basecode LIKE '%ICD10:E11%' OR
  c_basecode LIKE '%ICD10:E12%' OR
  c_basecode LIKE '%ICD10:E13%'
ORDER BY c_tooltip
"

tooltips <- dbGetQuery(c_connect, sql_explore_tooltip)
tooltips %>%
  unlist %>%
  cat(sep="\n")

```

Let's flip this around and find if there are any other concept_cd/c_basecode values associated with c_tooltip values containing diabetes codes.

```{r alternate-codes}
log_time()
sql_alternate_diabetes_codes <- "
SELECT DISTINCT c_basecode 
FROM blueheronmetadata.heron_terms
WHERE
  c_tooltip LIKE '%250 Diabetes mellitus%' OR
  c_tooltip LIKE '%E08-E13 Diabetes mellitus%'
"

alternate_codes <- dbGetQuery(c_connect, sql_alternate_diabetes_codes)
alternate_codes %>% 
  dim
alternate_codes %>%
  sample_n(20) %>%
  arrange(C_BASECODE)

alternate_codes %>%
  unlist %>%
  sub(":.*$", "", .) %>%
  table
```

```{r alternate-count}
log_time()

sql_alternate_select <- nest("WHERE concept_cd IN (", sql_alternate_diabetes_codes, ")")
sql_alternate_count <- nest(sql_count_patients, sql_alternate_select)
cat(sql_alternate_count)

dbGetQuery(c_connect, sql_alternate_count) %>%
  format(big.mark=",")
```

This snippet adds a date restriction.

```{r dates-1}
log_time()
sql_restrict_dates <- "
AND TO_CHAR(start_date, 'yyyy-mm-dd') > '2010-01-01'
"
```

Here's what the combination of all three strings looks like.

```{r dates-2}
log_time()
sql_count_date_restricted <- 
  nest(sql_count_diabetics, sql_restrict_dates)
cat(sql_count_date_restricted)
dbGetQuery(c_connect, sql_count_date_restricted)
```

This snippet restricts to patients to females only. This represents a self-join.

```{r females-1}
log_time()
sql_restrict_to_females <- "
AND patient_num IN (
  SELECT DISTINCT patient_num
  FROM blueherondata.observation_fact
  WHERE
    concept_cd = 'DEM|SEX:f'
)"
sql_count_female_diabetics <- nest(sql_count_date_restricted, sql_restrict_to_females)
cat(sql_count_female_diabetics)
```

Unfortunately, it is rather slow, so I do not recommend that you use this approach.

```{r females-2, eval=FALSE}
log_time()
dbGetQuery(c_connect, sql_combine)
```

Extracting all the patient numbers requires a different start to the SQL query.

```{r extract-1}
log_time()
sql_extract_patients <- "
SELECT patient_num, start_date AS pdate
  FROM blueherondata.observation_fact
"
```

This is what the combined query looks like.

```{r extract-2}
log_time()
sql_extract_diabetics <- nest(sql_extract_patients, sql_select_diabetes)
cat(sql_extract_diabetics)

patient_list <- dbGetQuery(c_connect, sql_extract_diabetics)
names(patient_list) <- tolower(names(patient_list))
patient_list %<>% arrange(patient_num)
head(patient_list)
dim(patient_list)
```

Notice that patients appear more often than once in the query. To get a distinct set of patients, you need to group by patient number and then choose only one record. The simplest choice is the record with the smallest start-date.

you can do this within R, but be careful, because some of the base R functions that you'd be tempted to use, such as unique or duplicated can be very inefficient with large data frames. The distinct or summarize functions in dplyr are a better choice.

```{r distinct-1}
log_time()
library(dplyr)
library(magrittr)
patient_list                   %>% 
  group_by(patient_num)        %>%
  summarize(pdate=min(pdate))  %>%
  ungroup                      -> unique_patients_1
head(unique_patients_1)
dim(unique_patients_1)
```

This method also works.

```{r distinct-2}
log_time()
library(dplyr)
library(magrittr)
patient_list                             %>% 
  arrange(patient_num, pdate)            %>%
  distinct(patient_num, .keep_all=TRUE)  %>%
  ungroup                                -> unique_patients_2
head(unique_patients_2)
dim(unique_patients_2)
```

You can also get the same effect using just SQL code.

```{r distinct-3}
log_time()
sql_extract_unique_patients <- "
SELECT patient_num, MIN(start_date) AS pdate
FROM blueherondata.observation_fact
"

sql_group <- "
GROUP BY patient_num
ORDER BY patient_num"
```

```{r distinct-4}
log_time()
sql_extract_unique_diabetics <- nest(sql_extract_unique_patients, sql_select_diabetes, sql_group)
cat(sql_extract_unique_diabetics)
unique_patients_3 <- dbGetQuery(c_connect, sql_extract_unique_diabetics)
head(unique_patients_3)
dim(unique_patients_3)
```

Suppose we want to know the gender of all these patients.

```{r gender-1, eval=FALSE}
log_time()
sql_extract_gender <- "
SELECT patient_num, concept_cd
FROM blueherondata.observation_fact
WHERE patient_num IN (
  SELECT DISTINCT patient_num 
  FROM blueherondata.observation_fact
  WHERE concept_cd LIKE '%E11.2%' 
  AND rownum < 100
) AND concept_cd LIKE 'DEM|SEX:%'"
gender_list <- dbGetQuery(c_connect, sql_extract_gender)
head(gender_list)
dim(gender_list)
table(gender_list$CONCEPT_CD)
```

Now, let's take things one step further. For these unique patients, let's pull any record associated with medications.

```{r medication}
log_time()
sql_extract_medications_prefix <- "
SELECT patient_num, concept_cd, start_date AS edate
FROM blueherondata.observation_fact

WHERE patient_num IN ("
  
sql_extract_medications_postfix <- "
) AND 

concept_cd IN (
SELECT DISTINCT concept_cd FROM blueherondata.concept_dimension
  WHERE concept_path LIKE '\\i2b2\\Medications\\%'
  ORDER BY concept_cd)"

sql_extract_medications <- paste(
  sql_extract_medications_prefix,
  sql_extract_unique_patients,
  sql_select_diabetes,
  sql_group,
  sql_extract_medications_postfix)
cat(sql_extract_medications)
```

Save everything for possible re-use

```{r save-everything}
log_time()
read.table(timing_file) 
save.image(image_file)
```
