---
title: "Apply exclusions to a patient list"
author: "Steve Simon"
date: "5/31/2018"
output: html_document
---

This program shows how to extract the patient numbers for diabetes patients.

Load the relevant libraries. 

```{r libraries}
library(dplyr)
library(knitr)
library(magrittr)
library("ROracle")
```

Please review open_oracle.Rmd to review how to access information in an Oracle database.

```{r timing-functions}
program_name <- "extract_diabetes_patient_numbers"
timing_file <- paste0("../src/", program_name, "_timing.txt")
image_file <- paste0("../data/", program_name, ".RData")

log_time <- function(first=FALSE) {
  current_chunk <- opts_current$get("label")
  current_time <- as.character(Sys.time())
  cat(current_time, current_chunk, "began.\n", file=timing_file, append=!first)
}
```


```{r login-to-oracle}
log_time(first=TRUE)
cdm_config <- read.csv('../cdm_config_B2.csv', stringsAsFactors=FALSE)
c_connect <-
  dbConnect(Oracle(), cdm_config$account, cdm_config$password, cdm_config$access)
```

Quick test

```{r quick-test}
dbGetQuery(c_connect, "SELECT * FROM blueherondata.observation_fact WHERE ROWNUM < 5")
dbGetQuery(c_connect, "SELECT * FROM blueherondata.concept_dimension WHERE ROWNUM < 5")
dbGetQuery(c_connect, "SELECT * FROM blueheronmetadata.heron_terms WHERE ROWNUM < 5")
```


The SQL code to count and extract patient numbers can get quite long. It helps to break the code into distinct pieces.

The following SQL snippet counts the distinct number of patients. You can combine with additional SQL code to place restrictions on the types of patients.

```{r count}
log_time()
sql_count_patients <- "
SELECT COUNT(DISTINCT patient_num)
FROM blueherondata.observation_fact
"

dbGetQuery(c_connect, sql_count_patients) %>%
  format(big.mark=",")
```

To make the SQL code look nicer, here is a function that indents a piece of SQL code nested between two other pieces.

```{r nest}
nest <- function(sql1, sql2, sql3="") {
  paste0(
    sql1,
    gsub("\n", "\n  ", sql2),
    sql3)
}
nestx <- function(sql1, sql2, sql3="", sql4="") {
  paste0(
    sql1,
    gsub("\n", "\n  ", sql2),
    gsub("\n", "\n    ", sql3),
    sql4)
}
```

This SQL snippet lists all the ICD9 and ICD10 codes associated with diabetes.

```{r diabetes-codes}
log_time()

sql_list_diabetes_icd9 <- "
SELECT DISTINCT concept_cd 
FROM blueherondata.concept_dimension
WHERE
  concept_cd LIKE '%ICD9:250%'
"

diabetes_codes <- dbGetQuery(c_connect, sql_list_diabetes_icd9)
diabetes_codes %>%
  dim
diabetes_codes %>%
  sample_n(5) %>%
  unlist %>%
  sort %>%
  cat(sep="\n")

sql_list_diabetes_icd10 <- "
SELECT DISTINCT concept_cd 
FROM blueherondata.concept_dimension
WHERE
  concept_cd LIKE '%ICD10:E08%' OR
  concept_cd LIKE '%ICD10:E09%' OR
  concept_cd LIKE '%ICD10:E10%' OR
  concept_cd LIKE '%ICD10:E11%' OR
  concept_cd LIKE '%ICD10:E12%' OR
  concept_cd LIKE '%ICD10:E13%'
"

diabetes_codes <- dbGetQuery(c_connect, sql_list_diabetes_icd10)
diabetes_codes %>%
  dim
diabetes_codes %>%
  sample_n(5) %>%
  unlist %>%
  sort %>%
  cat(sep="\n")
```

Combine these two queries together and nest them within a counting query.

```{r diabetes-count}
sql_list_diabetes_codes <- "
SELECT DISTINCT concept_cd 
FROM blueherondata.concept_dimension
WHERE
  concept_cd LIKE '%ICD9:250%' OR
  concept_cd LIKE '%ICD10:E08%' OR
  concept_cd LIKE '%ICD10:E09%' OR
  concept_cd LIKE '%ICD10:E10%' OR
  concept_cd LIKE '%ICD10:E11%' OR
  concept_cd LIKE '%ICD10:E12%' OR
  concept_cd LIKE '%ICD10:E13%'
"

sql_count_diabetics <- nestx(
  sql_count_patients, 
  "WHERE concept_cd IN (", 
  sql_list_diabetes_codes, ")")
cat(sql_count_diabetics)
dbGetQuery(c_connect, sql_count_diabetics) %>%
  format(big.mark=",")
```

Information about diabetes codes is also available through the concept_path variable found in the concept_dimension table.

```{r explore-path}
log_time()

sql_explore_icd9_path <- "
SELECT DISTINCT concept_path
FROM blueherondata.concept_dimension
WHERE
  concept_cd LIKE '%ICD9:250%'
ORDER BY concept_path
"

paths <- dbGetQuery(c_connect, sql_explore_icd9_path)
paths %>%
  dim
paths %>%
  sample_n(10) %>%
  unlist %>%
  sort %>%
  cat(sep="\n")

sql_explore_icd10_path <- "
SELECT DISTINCT concept_path
FROM blueherondata.concept_dimension
WHERE
  concept_cd LIKE '%ICD10:E08%' OR
  concept_cd LIKE '%ICD10:E09%' OR
  concept_cd LIKE '%ICD10:E10%' OR
  concept_cd LIKE '%ICD10:E11%' OR
  concept_cd LIKE '%ICD10:E12%' OR
  concept_cd LIKE '%ICD10:E13%'
ORDER BY concept_path
"

paths <- dbGetQuery(c_connect, sql_explore_icd10_path)
paths %>%
  dim
paths %>%
  sample_n(5) %>%
  unlist %>%
  sort %>%
  cat(sep="\n")
```

Unfortunately, the concept_path variable is mostly coded using cryptic values. If you stare at this list long enough, you'll figure out that the codes "A8352361" or "(250) Diabetes mellitus" will give you the proper ICD9 codes and that "A18916268" will give you the proper ICD10 codes 

You can find cleaner descriptions using c_tooltip in blueheronmetadata.heron_terms. The halibut system uses c_tooltip for tool tip displays, so this value is almost always easier to read.

```{r explore-tooltip}
sql_explore_icd9_tooltip <- "
SELECT DISTINCT c_tooltip 
FROM blueheronmetadata.heron_terms
WHERE
  c_basecode LIKE '%ICD9:250%'
ORDER BY c_tooltip
"

tooltips <- dbGetQuery(c_connect, sql_explore_icd9_tooltip)
tooltips %>%
  dim
tooltips %>%
  sample_n(5) %>%
  unlist %>%
  sort %>%
  cat(sep="\n")

sql_explore_icd10_tooltip <- "
SELECT DISTINCT c_tooltip 
FROM blueheronmetadata.heron_terms
WHERE
  c_basecode LIKE '%ICD10:E08%' OR
  c_basecode LIKE '%ICD10:E09%' OR
  c_basecode LIKE '%ICD10:E10%' OR
  c_basecode LIKE '%ICD10:E11%' OR
  c_basecode LIKE '%ICD10:E12%' OR
  c_basecode LIKE '%ICD10:E13%'
ORDER BY c_tooltip
"

tooltips <- dbGetQuery(c_connect, sql_explore_icd10_tooltip)
tooltips %>%
  dim
tooltips %>%
  sample_n(5) %>%
  unlist %>%
  sort %>%
  cat(sep="\n")
```

Very nice! For ICD9 codes associated with diabetes, use "250 Diabetes mellitus"" and for ICD10 codes associated with diabetes, use "E08-E13 Diabetes mellitus". You can find key phrases in the tool tip using Heron as well.

Let's flip this around and find if there are any other concept_cd/c_basecode values associated with c_tooltip values containing diabetes codes.

```{r alternate-codes}
log_time()
sql_alternate_icd9_codes <- "
SELECT DISTINCT c_basecode 
FROM blueheronmetadata.heron_terms
WHERE
  c_tooltip LIKE '%250 Diabetes mellitus%'
"

alternate_codes <- dbGetQuery(c_connect, sql_alternate_icd9_codes)
alternate_codes %>% 
  dim
alternate_codes %>%
  sample_n(5) %>%
  arrange(C_BASECODE)

sql_alternate_icd10_codes <- "
SELECT DISTINCT c_basecode 
FROM blueheronmetadata.heron_terms
WHERE
  c_tooltip LIKE '%E08-E13 Diabetes mellitus%'
"

alternate_codes <- dbGetQuery(c_connect, sql_alternate_icd10_codes)
alternate_codes %>% 
  dim
alternate_codes %>%
  sample_n(5) %>%
  arrange(C_BASECODE)
```

Combine these together to see if we have a larger count.

```{r alternate-count}
log_time()

sql_alternate_diabetes_codes <- "
SELECT DISTINCT c_basecode 
FROM blueheronmetadata.heron_terms
WHERE
  c_tooltip LIKE '%250 Diabetes mellitus%' OR
  c_tooltip LIKE '%E08-E13 Diabetes mellitus%'
"

sql_alternate_select <- nest("WHERE concept_cd IN (", sql_alternate_diabetes_codes, ")")
sql_alternate_count <- nest(sql_count_patients, sql_alternate_select)
cat(sql_alternate_count)

dbGetQuery(c_connect, sql_alternate_count) %>%
  format(big.mark=",")
```

Extracting all the patient numbers requires a different start to the SQL query.

```{r extract-patients}
log_time()
sql_extract_patients <- "
SELECT patient_num, start_date AS pdate
  FROM blueherondata.observation_fact
"
sql_extract_diabetics <- nest(sql_extract_patients, sql_alternate_select)
cat(sql_extract_diabetics)

patient_list <- dbGetQuery(c_connect, sql_extract_diabetics)
names(patient_list) %<>% tolower
patient_list %>%
  dim
patient_list %>%
  sample_n(5) %>%
  arrange(patient_num)
```

Notice that patients appear more often than once in the query. To get a distinct set of patients, you need to group by patient number and then choose only one record. The simplest choice is the record with the earliest start-date.

you can do this within R, but be careful, because some of the base R functions that you'd be tempted to use, such as unique or duplicated can be very inefficient with large data frames. The distinct or summarize functions in dplyr are a better choice.

```{r extract-distinct}
log_time()
library(dplyr)
library(magrittr)
patient_list                   %>% 
  group_by(patient_num)        %>%
  summarize(pdate=min(pdate))  %>%
  ungroup                      -> unique_patients_1
unique_patients_1 %>%
  dim
unique_patients_1 %>%
  sample_n(5) %>%
  arrange(patient_num)
```

This method also works.

```{r alternate-distinct}
log_time()
library(dplyr)
library(magrittr)
patient_list                             %>% 
  arrange(patient_num, pdate)            %>%
  distinct(patient_num, .keep_all=TRUE)  %>%
  ungroup                                -> unique_patients_2
unique_patients_2 %>%
  dim
unique_patients_2 %>%
  sample_n(5) %>%
  arrange(patient_num)
```

You can also get the same effect using just SQL code.

```{r distinct-using-sql}
log_time()
sql_extract_unique_patients <- "
SELECT patient_num, MIN(start_date) AS pdate
FROM blueherondata.observation_fact
"

sql_group <- "
GROUP BY patient_num
ORDER BY patient_num"

sql_extract_unique_diabetics <- nest(sql_extract_unique_patients, sql_alternate_select, sql_group)
cat(sql_extract_unique_diabetics)

unique_diabetics <- dbGetQuery(c_connect, sql_extract_unique_diabetics)
names(unique_diabetics) %<>% tolower
unique_diabetics %>%
  dim
unique_diabetics %>%
  sample_n(5) %>%
  arrange(patient_num)
```

Save this SQL code and this list. You'll need it later.

```{r save_distinct}
save(list=c("sql_extract_unique_diabetics", "unique_diabetics"), file="../data/diabetes_patients.RData")
```

Save everything for possible re-use

```{r save-everything}
log_time()
read.table(timing_file) 
save.image(image_file)
```
