---
title: "Opening an Oracle database"
author: "Steve Simon"
date: "5/31/2018"
output: html_document
---

This program shows how to open an Oracle database and count the number of records.

To open Oracle, you need to have a small text file stored with information about Oracle. For the purposes of this program, I have used a file called cdm_config.csv, but any name will do fine.

You can't peek at my file because it includes confidential information, but here is an example of what that file might look like.

account,password,access

GeorgeBrett,secret_password,//pointer_to_database_location

Make sure there is a line in the file .gitignore reading

cdm_config.csv

or your private password will be posted on a public repository.

```{r login-to-oracle}
library("ROracle")
cdm_config <- read.csv('../cdm_config_A1.csv', stringsAsFactors=FALSE)
c_connect <-
  dbConnect(Oracle(), cdm_config$account, cdm_config$password, cdm_config$access)
```

If your connection works then the following code will count the number of records in the observation_fact table. On my system, this produces a number in the low billions.

```{r simple-test}
n <- dbGetQuery(c_connect, "SELECT COUNT(patient_num) FROM blueherondata.observation_fact")
format(n, big.mark=",")
```

Recently, R Markdown has added features that allow you to run SQL code directly. Here's how that would work.

```{sql direct-access, connection=c_connect, output.var="total_count"}
SELECT COUNT(patient_num) FROM blueherondata.observation_fact
```

The output.var argument in the code above lets you tell the system where you want to results of the query to be stored in R.

```{r display-query}
format(total_count, big.mark=",")
```
